# Visualisation and communication {#viz}

Just as there are many types of data e.g. tables and
audio files, correspondingly there are many types of data visualisation e.g.
statistical plots and maps. Likewise visualisations we do for an exploration may
be different to those we do for communication. Therefore we narrow the scope here
to introduce some fundamentals of visualisation, presentations and reports.

By the end of this chapter you will be able to:

> + Use the `ggplot2` package to create exploratory plots and customize their plots
> + Use geometrical objects and aesthetic mappings to create box plots, bar plots and
time series plots
> + Use facets to split plots into sub-plots according to variables
> + Transform plots using statistical mappings and categorical variables (factors).
> + Transform data positions and coordinates
> + Use `ggplot2` themes and the `scales` package to customize plots
> + Export figures for use in documents and presentations
<!-- > + Create reports and presentations directly from RStudio -->

These skills will enable to start creating visualisations as well as
reports and presentations using R, 
but importantly it is up to 
you to determine what sort of visualisation is 
appropriate for the task at hand and for you to use your critical judgement in 
assessing a visualisation, presentation or report.

## Visualisation overview

For a deeper understanding of the art of data visualisation, a good place to start
is with the work of [Albert Cairo](http://www.thefunctionalart.com/), 
a journalist and academic specialising in data visualisation who has written and continues to write 
extensively on the subject. 

Betsy Mason has written an article that summarises many of the key ideas:
[Why scientists need to be better at data visualization](https://www.knowablemagazine.org/article/mind/2019/science-data-visualization)
including the issues around how we perceive shade and hue, that can make plots such
as heatmaps problematic.

Rafeal Irizarry has also written about [Why dynamite plots must die](https://simplystatistics.org/2019/02/21/dynamite-plots-must-die/). These
are very common plots in biomedical science, but are far from optimal.

The point here is not to tell you explicitly what plots to make, or that heatmaps
are bad etc., but to encourage you to think about what type of plot is best for the 
task in hand. Choosing to plot your data in a certain way because, *"that's what
everyone else does"* is unlikely to be the best reason.

To begin with two questions one might immediately have are:

1. Why visualise our data in the first place?
2. What do we know about what makes an effective visualisation?

### Making comparisons

Starting with the second question, a helpful set of principles for statistical 
visualisations is the work of Cleveland and McGill, namely:
[Graphical Perception: Theory, Experimentation, and Application to the Development of Graphical Methods by Cleveland and  McGill](https://www.jstor.org/stable/pdf/2288400.pdf?refreqid=excelsior%3Ab321aba183cce22a7f93335ca107eec6).

What they established is a hierarchy of perception, that is to say we there is an
order to which our brains find it easier or harder to make comparisons with visual information.

The order of comparisons from easiest to hardest is as follows, 
but this is best illustrated by some plotting some example data. 

1. Positions on a common scale
2. Positions on the same but non-aligned scales
3. Lengths
4. Angles, slopes
5. Area
6. Volume, colour saturation
7. Colour hue

Figure \@ref(fig:viz-plot) illustrates plotting the same percentages for five Countries seven 
ways corresponding with Cleveland and McGill's hierarchy.

```{r viz_dat, echo=TRUE, cache=F}
viz_dat <- tibble(Country =LETTERS[seq( from = 5, to = 1 )],
                  Percent = c(0.22,0.25,0.34,0.29,0.32)) 
```

```{r echo=FALSE, cache=F}

blank_theme <- theme_minimal() +
  theme(legend.position = "",
  axis.ticks = element_blank(),
              axis.text.x = element_blank(),
              axis.text.y = element_blank(),
              axis.title.x = element_blank(),
              axis.title.y = element_blank(),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  plot.title=element_text(size=14, face="bold")
  )

p1 <- viz_dat %>% ggplot(aes(Country,Percent)) +
        geom_bar(stat = "identity") + 
        coord_flip() 

p2 <- viz_dat %>% ggplot(aes(Country,Percent)) +
        geom_segment(aes(x=Country, xend=Country, y=0, yend=Percent)) +
        geom_point() +
        coord_flip()

p3 <- viz_dat %>% ggplot(aes(Country,Percent)) +
        geom_point() +
        coord_flip()

p4 <- viz_dat %>% ggplot(aes(Country, 1, size = Percent, label = Country)) +
        geom_point(colour = 'grey') +
        geom_text(hjust = 0.4, nudge_x = 0.35, size = 5) +
        blank_theme

p5 <- viz_dat %>% ggplot(aes(area =Percent,label = Country)) +
        geom_treemap() + 
        geom_treemap_text(colour = "white")

p6 <- viz_dat %>% ggplot(aes(x ="", y = Percent, fill = Country)) +
        geom_bar(stat = "identity") +
        coord_polar(theta = "y") +
        blank_theme +
        scale_fill_brewer(palette="Dark2") +
        geom_text(aes(label = Country), y = c(4.1,3.8,-2.15,1.8,2.95), colour = "white", size = 4)
      
p7 <- viz_dat %>% ggplot(aes(x =Country,y =1, fill= Percent)) +
  geom_tile() + blank_theme +
  geom_text(aes(label = Country), y = 1, colour = "white", size = 10)
```


(ref:viz-plot) The same percentages for five Countries A-E is plotted seven ways to illustrate the differences in ease of making comparisons depending the plot type. The percentages are A:32%, B:29%, C:34%, D:25% and E:22% in all plots.

```{r viz-plot,fig.cap='(ref:viz-plot)', echo=FALSE, cache=F}
plot_grid(p1,p2,p3,p4,p5,p6,p7)
```

The message here is that when choosing a visualisation for comparisons between 
sets of observations the best place to start is with plotting the observations on 
a common scale (but not necessarily finish).

## `ggplot2()`

In his paper [A layered grammar of graphics](http://vita.had.co.nz/papers/layered-grammar.pdf)
Hadley Wickham discusses the theory behind the `ggplot2()` package. As with tidyverse
the aim was to create a consistent system for building plots. 

The key components of a plot are:

+ data with aesthetic mappings e.g. position on an axis, colour or size.
+ geometric objects e.g. points, lines.

To this other components or layers can be added, such as facets or statistical transformations
to make increasingly complex plots.

As `ggplot2()` was first released in 2007 it preceded the `magrittr` package in 
2014 from which we get the pipe `%>%` it has one syntax difference that is sure
to catch you out. Whilst we can pipe into a `ggplot()` function, we build the plots using
the plus sign `+` to add our next layer. 

Likewise, forgetting to use the aesthetic function `aes()` within a layer is a common
mistake I make. Aesthetics and geometries are related, so for example you can't
set the shape aesthetic of a line. I usually make this mistake with colour and fill
aesthetics, where I try to use one when I need the other.

The basic form for creating plots with the `ggplot2` package is as follows:

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION(mapping = aes(<MAPPINGS>))>
```

To recap: 

> + We provide a data frame as an argument to the `ggplot()` function
> + Variables in the data frame are mapped to visual properties using the aesthetics function `aes()` 
> + The aesthetics are mapped to a geometric object such as points using a layer with a `geom()` function. 
> + If we were to plot several `geoms()` on the same plot wanted to map the same aesthetics to each `geom()` we can pass the aesthetics argument to `ggplot()` directly: `ggplot(data = <DATA>,mapping = aes(<MAPPINGS>))`.
> + We can also pipe data into `ggplot()` as we would for other functions.

### Datasaurus dozen

Returning to the first question: Why visualise data in the first place?

Inspired by the [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) and [Albert Cairo's Datasaurus](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html), Justin Matejka and
George Fitzmaurice
created the [Datasaurus dozen](https://www.autodeskresearch.com/publications/samestats) to 
illustrate a problem with relying only on summary statistics to understand data.

```{r datasaurus-dl, cache=TRUE}
# Download the Datasaurus Dozen
datasaurus <- read_tsv("https://raw.githubusercontent.com/ab604/coding-together/master/exercises/DatasaurusDozen.tsv")
# Quick look
glimpse(datasaurus)
# How many datasets
datasaurus %>% distinct(dataset)
```

We have a bakers' dozen of datasets, each with 1,846 observations for 
variables `x` and `y`. Let's look at the mean and standard deviation for each 
variable using a grouped summary and the `mean()` and standard deviation `sd()` functions. 

```{r datasaurus-summary, cache=TRUE}
datasaurus %>% group_by(dataset) %>% summarise(mean_x = mean(x),
                                               mean_y = mean(y),
                                               sd_x = sd(x),
                                               sd_y = sd(y))
```

Here we use the basic form to plot the `datasaurus` but with the addition of `facet_wrap()`
in which the first argument is `~` representing "formula" (a type of data structure) and the 
second argument is the variable we wish to create individual plots for. 
In words read this as *"facets depend upon the dataset"*:

```{r datasaurus-plot, cache=TRUE}
 ggplot(data = datasaurus) +
        geom_point(mapping = aes(x,y)) +
        facet_wrap(~ dataset)
```

The message here is that plotting the data reveals structure that is not apparent
from summary statistics. Therefore if possible: **plot early, and plot often.**

### Exploratory plots

As the name implies, the point of an exploratory plot is to explore our data. 
This can happen before, during or after tidying/transforming data, and as discussed in 
[The Art of Data Science](https://leanpub.com/artofdatascience) serves two key
purposes that we saw with the Datasaurus:

1. Creating expectations
2. Checking for deviations from our expectations.

In Chapter \@ref(data-wrangle-1) having transformed the Portal rodent surveys data 
into a table summarising the rodent observations per three month period per
plot type (control or exclosure). The original paper was exploring the hypothesis
that kangaroo rat and granivore populations were in competition for resources in their habitat.

We plotted the average number of captures over time as a line and point plot,
coloured according to plot type, and facetting the plot according to animal type.

Here we are exploring the effect of the exlosure on the animal populations over
time, and the variation in populations over time in both plot types.

(ref:by-q-tp) The average number of captures over time as a line and point plot, coloured according to plot type, and facetting the plot according to animal type.

```{r by-quarter-time-plot,fig.cap='(ref:by-q-tp)',fig.asp=1, out.width = '100%', fig.align='center',echo=TRUE,cache=TRUE, warning=FALSE}
ggplot(data = by_quarter,
       mapping = aes(x=quarter,y=mean_captures,colour=plot_type)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ rodent_type)
```

Expectations created by this plot for example are that the populations fluctuate
quite a lot over time, and the exclosure appears to work well with few kangaroo
rats observed in those plots.

What is less clear is what the effect of the exclosure is having, if any, on the
granivore population, suggesting we might wish to try a different type of plot.

### Statistical transformations

Often we want to transform the data as part of the plotting process, for example
to create plots that reveal statistical information such as distributions or 
averages. For these we can use geoms which are statistical in nature or to
which we provide statistical arguments.

#### Barplots

These simple plots actually reveal subtleties in 
plots. In the following code we take the `by_month_species` subset and create
a bar chart of `species_id`, filling the bars with colour according to `rodent_type`.

```{r,by-month-bar,purl=FALSE,echo=TRUE}
# Create a bar plot of genus from by_month_species 
by_month_species <- surveys_subset %>% 
        filter(rodent_type != "Other") %>% 
        group_by(year,month,species_id,rodent_type,plot_type) %>% 
        summarise(captures = n()/31)

by_month_species %>% 
  ggplot() +
  geom_bar(mapping = aes(x = species_id, fill=rodent_type)) + 
  facet_wrap(~ plot_type)
```


We see that the code has automatically created a count variable on the y-axis 
for each species_id plotted as a bar on the x-axis. 

This is because the `geom_bar()` algorithm automatically performs a statistical
transformation of the mapped variable. That is to say it performs a count and 
bins the data according to the genus. This means we have a chart where the bar
height corresponds to the number of rows for each species_id.

But this isn't a count of the captures, it's how many rows each species_id appears
in of `by_month_species`. Remember how we summarised this data to calculate captures per month.

Often when people talk about bar charts they are referring to when the height
of the bar is a variable present in the data, here that would be the `captures` 
column.
In this kind of situation we need to map the y-axis too, and change the
default `geom_bar()` statistical transformation to `stat = "identity"` so that
it takes the values provided in the `captures` column to determine the height of
the bar.

To reiterate: **if you are trying to create a bar chart using x and y variables
contained in your data set, you need to set `geom_bar(stat = "identity")`.** 

```{r,by-month-bar-stat,purl=FALSE,echo=TRUE}
# Create a bar plot of genus from by_month_species using y = captures and 
# stat = "identity" 
by_month_species %>% 
  ggplot() +
  geom_bar(mapping = aes(x = species_id, y = captures, fill=rodent_type), stat = "identity") + 
  facet_wrap(~ plot_type) 
```

Let's check this by looking at `PE` in the control plots and counting the rows and comparing that
to summing the number of captures. Note we need to `ungroup()` before doing another
summary. There are 132 rows, but only 11 captures.

```{r}
# How many rows for PE
by_month_species %>% 
  filter(species_id == "PE", plot_type == "Control") %>% ungroup() %>% tally()
# How many captures for PE
by_month_species %>% 
  filter(species_id == "PE", plot_type == "Control") %>% ungroup() %>% 
  summarise(total = sum(captures))
```

#### Boxplots

The previous exploratory plot showed us changes with time, but
averages and spread or other statistical summaries of the surveys data might be more informative
for understanding the effect of exlosure on granivore populations.

Box-plots are a standard way to plot the distribution of a data set. Plotting 
the data this way means that we drop the dynamic information, the time dimension,
but gain a compact summarised view of the variability of the number of captures,
and hopefully evidence of any changes in granivore population following exclosure.

In the following code we use the `boxplot()` geom, and this time swap the mapping
such that colour maps to the rodent type, and plots are faceted according to the plot type.

```{r,by_quarter_box_1,purl=FALSE,echo=TRUE}
ggplot(data= by_quarter,
       mapping = aes(x = rodent_type, 
                     y = mean_captures, 
                     colour = rodent_type)) +
  geom_boxplot() +
  facet_wrap(~ plot_type)
```

Comparing the median bar and the interquartile range, we can see evidence suggesting
a modest increase in the granivore population when kangaroo rats are excluded.

### Position adjustments

Another common adjustment we might make to our plots is to adjust the position 
of the data.

Recall the `mpg` data set for cars and our plot showing the fuel efficiency
versus the size of the engine. Many of the points are actually over plotted 
such that we can't see them.

By adding a bit of random noise to the position each data point, called `jitter`
we can spread the points out and see them more clearly. This is seen in the 
second plot below, where the `position = "jitter"` argument has been provided
to the `geom_point()` function.

It's a little counter intuitive to add noise to a plot, but when we're trying to
explore data for patterns it can be very useful.

```{r,mpg_jitter,purl=FALSE,echo=TRUE}
# Without jitter
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class))
# With jitter
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class),
             position = "jitter")
```

Let's add points to our box-plot and try comparing the plot
to one where the points are jittered.

```{r,by_quarter_box_2,purl=FALSE,echo=TRUE}
# Boxplot with points and jitter
ggplot(data= by_quarter,
       mapping = aes(x = rodent_type, y = mean_captures, 
                     colour = rodent_type)) +
  geom_boxplot() +
  geom_point(position = "jitter", alpha = 0.5) +
  facet_wrap(~ plot_type)
```

We can also apply positional adjustments to other plots such as barplots.
We can combine aesthetic `fill = species_id` with `position = "fill"` in 
`geom_bar()` to create a separate bar for each species and put plot_type on 
the x axis for `by_month_species` plots bars of length 1, but filled
with the proportion of captures for that species.

```{r,by-month-bar-dodge,purl=FALSE,echo=TRUE}
# Create a bar plot of genus from by_month_species using x = plot_id and 
# stat = "identity" 
by_month_species %>%
  ggplot() +
  geom_bar(mapping = aes(x = plot_type, y = captures,  fill=species_id), 
           position = "fill" , stat = "identity") 
```

Perhaps not that useful given what we know about making comparisons.

For more positional adjustments, check out the help files `?postion_jitter`,
`?position_dodge`, `?position_fill`, `?position_identity`,  and 
`?position_stack`.

### Coordinate adjustments

Previously, we plotted `by_month_species` and sometimes it's tricky when there are
lots of labels on the x-axis, so let's flip the plot round using `coord_flip()` to improve things:

```{r,by-month-bar-5,purl=FALSE,echo=TRUE}
# Flip the coordinates to make the labels more readable 
ggplot(by_month_species) +
  geom_bar(mapping = aes(x = species_id, y = captures, fill = rodent_type), 
           stat = "identity") + 
  facet_grid(~ plot_type) +
  coord_flip()
```

#### Factors

Finally for this plot, it would be nice to put the bars in order of size. We
can do this by converting the genus to a factor.

Recall that factors are how R represents categorical variables, variables with
a limited number of values, such as genus.

Also recall **factors look like strings, but behave like integers**. That is to 
say they are strings of characters with associated values called levels 
that can be used to order catergories.

This is why factors are useful, especially when we want to place things in 
non-alphabetical order. We can take advantage of factor levels to order things
as we wish.

Check out the [forcats](http://forcats.tidyverse.org/) tidyverse package to 
explore the power of factors, but here to give you a taste of what is possible
we're going to use the `forcats` package `fct_reorder()` function to convert 
`species_id` into a factor and in doing so set the level according to to the number 
of captures. In this way we can order the bars in our plot according to the 
number of captures and see the pattern more clearly.

In `fct_reorder()`, the first argument is the variable we wish to make into a 
factor `species_id`, and the second argument is variable we wish to use to create 
the order. Here we want to order according to the number of captures, 
so `captures` is the second argument. 

Note we still keep x and y mappings, but now x is now a function of both the 
genus and the capture columns.

This gives us a nice ordered bar chart by genus and number of captures during 
the experimental period. 

**We can clearly see that kangaroo rat exclosure works,
and that this corresponds with an increase in granivore captures these plots.**

```{r,by_month_bar_6,purl=FALSE,echo=TRUE}
# Load the forcats tidyverse package
library(forcats)

# Order the rodents according to number of captures
ggplot(by_month_species) +
  geom_bar(mapping = aes(x = fct_reorder(species_id, captures), y = captures, 
                          fill = rodent_type), stat = "identity") + 
  facet_grid(~ plot_type) +
  coord_flip()
```

## Themes and customisations

When we're exploring our data we generally don't mind if the labels and colours
etc... are a bit messy. But when we want to publish or communicate our findings
to others we want everything to be just so.

This is where the code can become quite complicated and it's hard to provide
a general case. However, the underlying template remains the same:

```{r ggplot-full-template,,eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>
```

It's unlikely that you'll need to statistical transformations, positional 
adjustments, coordinate adjustments, and facets all in the same plot, but this 
provides an idea of what is possible. Remember the aim is clarity, not to
make something complicated just because you can.

This template we can use to add all the additional functions and arguments that 
will change the plot to make it exactly how we want. It takes time, but remember 
code is reusable. Solve it once and you've solved it forever.

For all the possible customisations check out the 
[ggplot2 documentation](http://ggplot2.tidyverse.org/reference/index.html), and
this chapter of R for data science: 
[Graphics for communication](http://r4ds.had.co.nz/graphics-for-communication.html#themes)

But here's a few common changes.

### Changing colours, labels, themes and adding titles

Recall our box-plot with points overlaid. Following the facet function, 
we're going add a succession of functions to change the colours 
`scale_colour_brewer()`, change the axis labels, 
add a title using `labs()` and change the theme.

Themes deal with non-data elements e.g. the grid, the axes etc. There are
a range of built-in themes and packages of themes such as `ggthemes`. Here
we'll use the built-in theme, `theme_bw()` that gives a simple layout.

```{r,by-quarter-box4,purl=FALSE,echo=TRUE}

ggplot(data= by_quarter,
       mapping = aes(x = rodent_type, y = mean_captures, 
                     colour = rodent_type)) +
  geom_boxplot() +
  geom_point(position = "jitter") +
  facet_wrap(~ plot_type) +
  scale_colour_brewer(palette = "Set1") +
  labs(title = "Quarterly average captures of rodents between 1977 and 1990",
       x = "Rodent type",
       y = "Number of captures",
       colour = "Rodent type") +
  theme_bw()

```

Check out `?scale_colour_brewer` and `?scale_fill_brewer` for more information 
on colour schemes for discrete data. And `?scale_color_continuous` for creating colour gradients for continuous data. 

<!-- ## Reports -->

<!-- Using R for report writing and presentations. -->

<!-- ## Presentations -->

<!-- My personal go-to for slide design guidance is [Melinda Seckington](https://missgeeky.com/2017/08/04/the-art-of-slide-design/). She -->
<!-- has written a fantastic step-by-step guide to creating effective slides. -->

<!-- Condensing her advice down she suggests: -->

<!-- > + Focus on one point per slide -->
<!-- > + Don't distract -->
<!-- > + Make important information stand out using what we know about graphical perception -->
<!-- > + Be consistent -->

<!-- I created a short [example presentation about cleaning your teeth](https://docs.google.com/presentation/d/e/2PACX-1vTa8VfB_jpB6fG4i157eNJb8sAyyKjJW06iZ8NNL1LPcB40w4NLLpK3TetbCuWSckX3wc8XuqdOpfnf/pub?start=false&loop=false&delayms=30000) -->
<!-- following these ideas. -->
