# Visualistion and communication {#viz}

Just as there are many types of data e.g. tables and
audio files, correspondingly there are many types of data visualisation e.g.
statistical plots and maps. Likewise visualisations we do for an exploration may
be different to those we do for communication. Therefore we narrow the scope here
to introduce some fundamentals of visualisation, presentations and reports.

By the end of this chapter you will be able to:

> + Use the `ggplot2` package to create exploratory plots and customize their plots
> + Use geometrical objects and aesthetic mappings to create box plots, bar plots and
time series plots
> + Use facets to split plots into subplots according to variables
> + Transform plots using statistcal mappings and catergorical variables (factors).
> + Transform data positions and coordinates
> + Use `ggplot2` themes and the `scales` package to customize plots
> + Export figures for use in documents and presentations
> + Create reports and presentations directly from RStudio

These skills will enable to start creating visualisations as well as
reports and presentations using R, 
but importantly it is up to 
you to determine what sort of visualiation is 
appropriate for the task at hand and for you to use your critical judgement in 
assessing a visualisation, presentation or report.

## Visualistion

[Albert Cairo](http://www.thefunctionalart.com/) is a journalist and academic
specialising in data visualisation and has written and continues to write 
extensively on the subject.

[Graphical Perception: Theory, Experimentation, and Application to the Development of Graphical Methods by Cleveland and  McGill](https://www.jstor.org/stable/pdf/2288400.pdf?refreqid=excelsior%3Ab321aba183cce22a7f93335ca107eec6)

### Datasaurus dozen

But why visualise data in the first place?

Inspired by the [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) and [Albert Cairo's Datasaurus](http://www.thefunctionalart.com/2016/08/download-datasaurus-never-trust-summary.html), Justin Matejka and
George Fitzmaurice
created the [Datasaurus dozen](https://www.autodeskresearch.com/publications/samestats) to 
illustrate a problem with relying only on summary statistics to understand data.

```{r datasaurus-dl, cache=TRUE}
# Download the Datasaurus Dozen
datasaurus <- read_tsv("https://raw.githubusercontent.com/ab604/coding-together/master/exercises/DatasaurusDozen.tsv")
# Quick look
glimpse(datasaurus)
# How many datasets
datasaurus %>% distinct(dataset)
```

```{r datasaurus-summary, cache=TRUE}
datasaurus %>% group_by(dataset) %>% summarise(mean_x = mean(x),
                                               mean_y = mean(y),
                                               sd_x = sd(x),
                                               sd_y = sd(y))
```

```{r datasaurus-plot, cache=TRUE}
datasaurus %>% 
        ggplot(aes(x,y)) +
        geom_point() +
        facet_wrap(~ dataset)
```


### Exploratory plots

### Polished plots

## Reports

Using R for report writing and presentations.

## Presentations

My personal go-to for slide design guidance is [Melinda Seckington](https://missgeeky.com/2017/08/04/the-art-of-slide-design/). She
has written a fantastic step-by-step guide to creating effective slides.

Condensing her advice down she suggests:

> + Focus on one point per slide
> + Don't distract
> + Make important information stand out using what we know about graphical perception
> + Be consistent

I created a short [example presentation about cleaning your teeth](https://docs.google.com/presentation/d/1ssXZYkuLzz9bXLeqmHmT66fpj7dpil0aQCY3yp36adA/edit?usp=sharing)
following these ideas.
